---
title: "Growth & Development in Primates"
author: "Kristen Tuosto"
date: "2023-08-14"
output: html_document
---

# Descriptive Statistics 

```{r csv files}

library(tidyverse)
library(ggplot2)
library(Hmisc)
library(dplyr)
library(forcats)

# Load data
df_tissue <- read.csv("C:/Users/ktuos/Downloads/df_wide_tissue.csv")

df_ID <- read.csv("C:/Users/ktuos/OneDrive/Desktop/Tissue_ID.csv")

# Merge data

df  <- merge(df_ID, df_tissue, by = 'ID')

```

# Examine structure of df
```{r}
# Check df structure 

str(df)

# Convert GENUS, SEX, and AGE into factors

df$GENUS <- factor(df$GENUS, level = c("chlorocebus", "hylobates", "pan", "gorilla"))

df$SEX <- as.factor(df$SEX)
df$AGE <- factor(df$AGE, levels = c("DECID", "M1", "M2", "M3", "EPIPH"))

# Drop pixel count variables and Percent.C.Ar

df2 <- df[,-5:-19]

```

# Check df structure again
```{r}

# Check df structure 

str(df2)

```

# Descriptive statistics by GENUS
```{r}
by(df2, df$GENUS, summary)
```

# Descriptive statistics by GENUS and AGE
```{r}
# List of variables you want to summarize
vars_to_summarize <- colnames(df2[, sapply(df2, is.numeric)])

# Summarize by Genus and AGE
df_summary <- df2 %>%
  group_by(GENUS, AGE) %>%
  summarise_at(
    .vars = vars_to_summarize,
    .funs = list(mean = ~mean(., na.rm = TRUE),
                 sd = ~sd(., na.rm = TRUE),
                 min = ~min(., na.rm = TRUE),
                 max = ~max(., na.rm = TRUE),
                 count = ~sum(!is.na(.)))
  )

df_summary
```

# Descriptive statistics by GENUS, AGE, and SEX
```{r}
# Summarize by Genus, AGE, and SEX
df_summary2 <- df2 %>%
  group_by(GENUS, AGE, SEX) %>%
  summarise_at(
    .vars = vars_to_summarize,
    .funs = list(mean = ~mean(., na.rm = TRUE),
                 sd = ~sd(., na.rm = TRUE),
                 min = ~min(., na.rm = TRUE),
                 max = ~max(., na.rm = TRUE),
                 count = ~sum(!is.na(.)))
  )

df_summary2
```
# Contingency table 
```{r}
# GENUS
table(df2$GENUS)

# GENUS & AGE
table(df2$GENUS, df2$AGE)

# GENUS & SEX 

table(df2$GENUS, df2$SEX)

# GENUS, AGE, & SEX
table(df2$GENUS, df2$AGE, df2$SEX)
```

# Sub setting data by cortical area
```{r}
# Sub-cortical area dfs

# Percent tissue type of cortical area df
df_c.ar <- df2[,-18:-27]

# Long format 
df_c <- df_c.ar %>% 
  gather(key = "Tissue", value = "Amount", -ID, -GENUS, -AGE, -SEX)

# Add two new columns with biological knowledge (cortex location and deposition speed)

# Cortex location 

df_c$Category <- sapply(df_c$Tissue, function(tissue) {
  if (tissue %in% c("C.Ar_ECCC", "C.Ar_ELAM", "C.Ar_EPF", "C.Ar_EPF.LAM")) {
    return("Endosteal")
  } else if (tissue %in% c("C.Ar_SF", "C.Ar_HAV", "C.Ar_INT")) {
    return("Cortex")
  } else {
    return("Periosteal")
  }
})

# Deposition speed

df_c$Speed <- sapply(df_c$Tissue, function(tissue) {
  if (tissue %in% c("C.Ar_SF", "C.Ar_HAV", "C.Ar_INT", "C.Ar_ECCC")) {
    return("Neutral")
  } else if (tissue %in% c("C.Ar_ELAM", "C.Ar_LAM")) {
    return("Slow")
  } else if (tissue %in% c("C.Ar_WOV", "C.Ar_FLC")) {
    return("Fast")
  } else {
    return("Moderate")
  }
})

# Rename tissue names 
df_c$Tissue <- recode(df_c$Tissue, 
                      `C.Ar_WOV` = "WOV",
                      `C.Ar_FLC` = "FLC",
                      `C.Ar_FLC.LZPO` = "FLC-LZPO",
                      `C.Ar_PF` = "PF",
                      `C.Ar_PF.LAM` = "PF-LAM",
                      `C.Ar_LAM` = "LAM",
                      `C.Ar_ECCC` = "CCC",
                      `C.Ar_EPF` = "EPF",
                      `C.Ar_EPF.LAM` = "EPF-LAM",
                      `C.Ar_ELAM` = "ELAM",
                      `C.Ar_INT` = "INT",
                      `C.Ar_SF` = "SF",
                      `C.Ar_HAV` = "HAV",
                      )


# Reordering tissue types by category and speed

# Ensure the columns are factors
df_c$Tissue <- factor(df_c$Tissue, levels = c("WOV", "FLC", "FLC-LZPO", "PF", 
                      "PF-LAM", "LAM", "EPF", "EPF-LAM", "ELAM", "CCC", "SF", "INT", "HAV"))



# Percent periosteal tissue type
  
df_p.ar <- df2[, c(1:4, 18:23)]

# Long format 
df_p <- df_p.ar %>% 
  gather(key = "Tissue", value = "Amount", -ID, -GENUS, -AGE, -SEX)

# Add one new column with biological knowledge (deposition speed)

# Deposition speed

df_p$Speed <- sapply(df_p$Tissue, function(tissue) {
  if (tissue %in% c("P.Ar_WOV", "P.Ar_FLC")) {
    return("Fast")
  } else if (tissue %in% c("P.Ar_LAM")) {
    return("Slow")
  } else {
    return("Moderate")
  }
})


# Rename tissue names 
df_p$Tissue <- recode(df_p$Tissue, 
                      `P.Ar_WOV` = "WOV",
                      `P.Ar_FLC` = "FLC",
                      `P.Ar_FLC.LZPO` = "FLC-LZPO",
                      `P.Ar_PF` = "PF",
                      `P.Ar_PF.LAM` = "PF-LAM",
                      `P.Ar_LAM` = "LAM"
                      )


# Reordering tissue types by category and speed

# Ensure the columns are factors
df_p$Tissue <- factor(df_p$Tissue, levels = c("WOV", "FLC", "FLC-LZPO", "PF", 
                      "PF-LAM", "LAM"))


# Percent endosteal tissue type
  
df_e.ar <- df2[,-5:-23 ]

# Long format 
df_e <- df_e.ar %>% 
  gather(key = "Tissue", value = "Amount", -ID, -GENUS, -AGE, -SEX)

# Add one new column with biological knowledge (deposition speed)

# Deposition speed

df_e$Speed <- sapply(df_e$Tissue, function(tissue) {
  if (tissue %in% c("E.Ar_LAM")) {
    return("Slow")
  } else if (tissue %in% c("E.Ar_ECCC")) {
    return("Neutral")
  } else {
    return("Moderate")
  }
})

# Rename tissue names 
df_e$Tissue <- recode(df_e$Tissue, 
                      `E.Ar_ECCC` = "CCC",
                      `E.Ar_EPF` = "EPF",
                      `E.Ar_EPF.LAM` = "EPF-LAM",
                      `E.Ar_ELAM` = "ELAM"
                      )

# Reordering tissue types by speed

# Ensure the columns are factors
df_e$Tissue <- factor(df_e$Tissue, levels = c("EPF", "EPF-LAM", "ELAM", "CCC"))

```

### Color palettes for plots
```{r}
# Tissue type color palette - all tissue 
color_palette <- c("darkgreen", "forestgreen", "mediumseagreen", "seagreen3", "aquamarine3", "aquamarine2", "indianred4", "indianred", "indianred2", "darkgoldenrod1", "lemonchiffon2", "darkseagreen1", "darkslateblue")

# Tissue type color palette - periosteal tissue
color_palette1 <- c("darkgreen", "forestgreen", "mediumseagreen", "seagreen3", "aquamarine3", "aquamarine2")

# Tissue type color palette - endosteal tissue 
color_palette2 <- c("indianred4", "indianred", "indianred2", "darkgoldenrod1")
```

```{r}
# Bar plot
ggplot(df_c, aes(GENUS, Amount, fill = Tissue), stat="identity", position = "dodge") + 
  stat_summary(fun.y = mean, geom = "bar", position = "dodge") + 
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", position = position_dodge(width = .96), width = 0.02) +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ylim(0, 80) +
  scale_fill_manual(values= color_palette)  

# The Amounts are based on means for each primate
```


# Plot composition of tissues relative to each other GENUS only consideration
```{r}

# Calculate the composition of tissues relative to each other
df_c_proportions <- df_c %>%
  group_by(GENUS) %>%
  mutate(Proportion = Amount / sum(Amount, na.rm = TRUE))

# The data represents the relative composition of tissues in terms of their contribution to the entire cortical area (i.e., the combined tissue percentages equal 100%).

# Plot composition of tissues relative to each other
ggplot(df_c_proportions, aes(x = GENUS, y = Proportion, fill = Tissue)) +
  geom_bar(stat = "identity", position = "stack", width = 0.90) + 
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(y = "Proportion", x = "Primate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), aspect.ratio = 5/10) +
  scale_fill_manual(values= color_palette) 

```

# All cortical bone tissue types bar plot
```{r}

# Bar plot
ggplot(df_c, aes(GENUS, Amount, fill = Tissue), stat="identity", position = "dodge") + 
  stat_summary(fun.y = mean, geom = "bar", position = "dodge") + 
  facet_wrap(~ AGE) + 
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", position = position_dodge(width = .96), width = 0.02) +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ylim(0, 80) +
  scale_fill_manual(values= color_palette)  

```

# Plot composition of tissues relative to each other
```{r}

# Calculate the composition of tissues relative to each other
df_c_proportions <- df_c %>%
  group_by(GENUS, AGE) %>%
  mutate(Proportion = Amount / sum(Amount, na.rm = TRUE))

# The data represents the relative composition of tissues in terms of their contribution to the entire cortical area (i.e., the combined tissue percentages equal 100%).

# Plot composition of tissues relative to each other
ggplot(df_c_proportions, aes(x = GENUS, y = Proportion, fill = Tissue)) +
  geom_bar(stat = "identity", position = "stack", width = 0.90) + 
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(y = "Proportion", x = "Primate") +
  facet_wrap(~ AGE, ncol = 1) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), aspect.ratio = 5/10) +
  scale_fill_manual(values= color_palette) 

```


# Periosteal tissue types bar plot
```{r}

# Bar plot
ggplot(df_p, aes(GENUS, Amount, fill = Tissue), stat="identity", position = "dodge") + 
  stat_summary(fun.y = mean, geom = "bar", position = "dodge") + 
  facet_wrap(~ AGE) + 
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", position = position_dodge(width = .96), width = 0.02) +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ylim(0, 80) +
  scale_fill_manual(values= color_palette1)  

```

# Plot composition of tissues relative to each other for periosteal area 
```{r}

# Calculate the composition of tissues relative to each other
df_p_proportions <- df_p %>%
  group_by(GENUS, AGE) %>%
  mutate(Proportion = Amount / sum(Amount, na.rm = TRUE))

# Plot composition of tissues relative to each other
ggplot(df_p_proportions, aes(x = GENUS, y = Proportion, fill = Tissue)) +
  geom_bar(stat = "identity", position = "stack", width = 0.90) + 
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(y = "Proportion", x = "Primate") +
  facet_wrap(~ AGE, ncol = 1) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), aspect.ratio = 5/10) +
  scale_fill_manual(values= color_palette1) 

```

# Endosteal bone tissue types bar plot
```{r}

# Bar plot
ggplot(df_e, aes(GENUS, Amount, fill = Tissue), stat="identity", position = "dodge") + 
  stat_summary(fun.y = mean, geom = "bar", position = "dodge") + 
  facet_wrap(~ AGE) + 
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", position = position_dodge(width = .96), width = 0.02) +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ylim(0, 80) +
  scale_fill_manual(values= color_palette2)  

```

# Plot composition of tissues relative to each other for the endosteal area 
```{r}

# Calculate the composition of tissues relative to each other 
df_e_proportions <- df_e %>%
  group_by(GENUS, AGE) %>%
  mutate(Proportion = Amount / sum(Amount, na.rm = TRUE))

# Plot composition of tissues relative to each other
ggplot(df_e_proportions, aes(x = GENUS, y = Proportion, fill = Tissue)) +
  geom_bar(stat = "identity", position = "stack", width = 0.90) + 
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(y = "Proportion", x = "Primate") +
  facet_wrap(~ AGE, ncol = 1) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), aspect.ratio = 5/10) +
  scale_fill_manual(values= color_palette2) 

```


